---
title: "Google Data Analytics Capstone: New car sales 2019-2021. How COVID impacted the industry?"
author: "Boris Kosarenkov"
date: "2022-10-17"
output: html_document
---

```{r, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r setup, include=FALSE}
install.packages('tidyverse')
library(tidyverse)
install.packages('DT')
library(DT)
install.packages('tidyr')
library(tidyr)
install.packages('scales')
library(scales)
install.packages("devtools")
library(devtools)
install_github("bbc/bbplot")
library(bbplot)
```

```{r upload data, include=FALSE}
sales <- read_csv('car_truck_sales_2019-2021_csv.csv')
sales_to_gdp <- read_csv ('EU passenger car sales vs GDP_v2.csv')
head(sales)
head(sales_to_gdp)
```

```{r DT usage, include=FALSE}
dt_sales <- datatable(sales, filter = c("top"))
dt_GDP_sales <- datatable(sales_to_gdp, filter = c("top"))
```

```{r pivoting longer, include=FALSE}
sales_long <- pivot_longer (
data = sales, 
cols = starts_with("ytd"),
names_to = "period",
names_prefix = "ytd_",
values_to = "sales"
)
head(sales_long)
```

```{r charts grouped, include=FALSE}
sales_groupped <- ggplot(data = sales_long) +
  geom_col(mapping = aes(x = period, y=sales, fill = as.factor(regions)), position="dodge") +
  scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6))+
  geom_hline(yintercept = 0, size = 1, colour="#333333")+
  scale_fill_brewer(palette = "Pastel1")+
  bbc_style()+
  labs(title = "New cars sales per region",
       subtitle = "Millions of cars sold per region 2019 - 2021", caption="Source: OICA")+
  theme(legend.position = "top", 
        legend.justification = "left", plot.caption = element_text(size=8))
```

```{r charts stacked, include=FALSE}
sales_stacked <- ggplot(data = sales_long) +
  geom_col(mapping = aes(x = period, y=sales, fill = regions)) +
  scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6))+
  geom_hline(yintercept = 0, size = 1, colour="#333333")+
  scale_fill_brewer(palette = "Pastel1")+
  bbc_style()+
  labs(title = "New cars sales per year",
       subtitle = "Millions of cars sold in 2019 - 2021", caption="Source: OICA")+
  theme(legend.position = "top", 
        legend.justification = "left", plot.caption = element_text(size=8))
```

```{r building sales_to_GDP chart, include=FALSE}
sales_to_gdp$Period = as.numeric(sales_to_gdp$Period)
sales_vs_GDP <- ggplot(sales_to_gdp, aes(x=Period))+
  geom_line(aes(y=Sales, colour = 'Sales'), size = 1.5)+
  geom_line(aes(y=GDP_annual_growth, colour = 'GDP_annual_growth'), size = 1.5)+
  scale_y_continuous(labels = label_number(suffix = " %"))+
  scale_x_continuous(breaks = sales_to_gdp$Period)+
  geom_hline(yintercept = 0, size = 1, colour="#333333")+
  bbc_style()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), legend.position = "top", 
        legend.justification = "left", plot.caption = element_text(size=8))+
  labs(title = "Car Sales vs GDP, Europe",
       subtitle = "GDP and car sales annual variation", caption = "Source: OICA, World Bank, ACEA")+
  scale_fill_brewer(palette = "Pastel1")
```

## Ask

In this capstone project, I will explore the new car sales and its COVID impact. My target is to understand when the industry will be able to come back to the pre-COVID level of 2019. Stakeholders are Global Supply Chain Management of one of the major automotive parts manufacturers. The business task is to analyze the automotive market and help stakeholders perform demand arbitration and inventory strategy for 2023.

## Prepare

I will use data from "Organisation Internationale des Constructeurs d'Automobiles" (OICA), The European Automobile Manufacturers' Association, or ACEA, and The World Bank. This paper is created for educational, research purposes, which falls under exceptions to French copyright law, described in Art. L122-5. The World Bank strives to enhance public access to and use of data that it collects and publishes ([detailed terms of use](https://www.worldbank.org/en/about/legal/terms-of-use-for-datasets))

-   Data-sets & citations:

    -   Global Sales - All Vehicles from OICA, [OICA web-site](www.oica.net)

    -   Vehicle sales mirror economic growth (2008-2021 trend), [ACEA web-site](www.acea.auto)

    -   The Wrold Bank: GDP growth (annual %),[The World Bank web-site](data.worldbank.org)

## Process

For Data processing, cleaning and analyze I will use Excel and R studio cloud. Excel is chosen as it fits very well for initial review of the data and making assumptions. I have significant experience with Excel, so it is faster and easier for me to make a quick initial data review. R programming language and R studio IDE is chosen because it has great capabilities in both data processing, transformation, and visualization. R markdown format allows to output professional reports in HTML. Same analyze and visualization could have been performed fully in Excel, or in Google Big Query (using SQL) and Google data studio or Tableu for visualization.

Technical documentation logging cleaning and manipulation of data is available on [Kaggle](https://www.kaggle.com/code/boriskosarenkov/google-data-analytics-car-sales-and-covid-impact)

## Analyze

I have prepared two data-sets. I used excel to delete and or modify column names, delete unnecessary visual formatting, and to save data sets as CSV files. In the next step I'm using R to load those CSV files into IDE. I use tidyverse package to manipulate data in R, I had to pivot longer one of the data sets in order to be able to build the required visualization. I use DT package to make the advanced visual output of the resulting tables. For charts, I use ggplot2 with a style package created by [BBC bbplot package](https://github.com/bbc/bbplot).

First data set: Global Sales - All Vehicles

-   It contains new vehicle sales per country and region for the period 2019 - 2021

```{r echo=FALSE}
dt_sales
```

-   It allows me (after some pivoting) to build a column chart representing the drop in sales in 2020

```{r echo=FALSE}
    sales_stacked
```

Second data set: GDP growth correlation with new vehicle sales

-   It was made using data from The World Bank for GDP annual growth and data from OICA for vehicle sales. An article from ACEA provided the insight that correlation exists

```{r echo=FALSE}
dt_GDP_sales
```

-   Line chart helps to demonstrate the assumed correlation

```{r echo=FALSE}
    sales_vs_GDP
```

## Act

-   Data shows almost 24% drop in new car sales in 2020 comparing to 2019. The reduction is related to COVID pandemic impact on customer demand and on the whole automotive supply chain. Car manufacturers and component suppliers suffered production shutdowns due to quarantines.
-   Data shows that there is an above average correlation between car sales and GDP growth. See chart.\
-   On the same chart we can see that in 2020 car sales "overreact" - drop significantly lower in comparison to the decline in actual economic emissions (GDP). "Overreaction" can be caused by:\
    -   Panic of consumers\

    -   Showrooms closed\

    -   Manufacturer's production output reduced due to crisis in supply chain\
-   World Bank estimates only 3.5% annual growth of GDP in 2023 and in 2024. With such pace market will comeback top the 2019 (pre-COVID) levels on in 2024

If stakeholders get a demand forecast from car makers (their customers) which contradicts above described insights, and for example shows sales growth above 3-4% next year it should raise questions for further analysis:

-   Is this carmaker or vehicle model outpacing the market? Did it ever before?

-   Is the region model is targeted for outpacing the market? For example, China is and it is estimated that it will in 2023 as well.

If none of the above is true I suggest stakeholders to be conservative in their demand arbitration process for 2023, so they could avoid overstocking.